Microplatforms: Product platforms as code

### 2018-03-14 paper.tex
more biblio

Starting with the software itself, it is crucial that it is thoroughly tested before any part of it can be considered done. Testing is now typically largely automated~\cite{FowlerTestPyramid}, although manual exploratory testing is also strongly advised [http://www.methodsandtools.com/archive/archive.php?id=65]. Testing activities fit on a scale of the level of integration of all the components involved, from testing each in isolation (e.g. unit testing) to fully integrated testing of the run-time behaviour of the entire application stack.

Typically, this is supported by a number of different environments [https://www.phparch.com/2009/07/professional-programming-dtap-\%E2\%80\%93-part-1-what-is-dtap/] starting from the lower ones, where external dependencies of services are stubbed [http://xunitpatterns.com/Test\%20Stub.html] and moving all the way up to the production environment which serves customers. Traditionally, these environments are named, long-lived, and are often reused for the purposes of various different testing activities (e.g. functional, performance, and penetration testing).

In order to get applications deployed, configured and started in these environments a process of deployment activities is followed, ranging from completely manual to fully automated. There are several different methodologies addressing the change-related risk of deploying to the production environment, ranging from large regular deployments done every few months all the way to Continuous Delivery and Continuous Deployment [https://continuousdelivery.com/].

@@

Typically, when following the continuous deployment method, businesses do not want customers to immediately see the change caused by every deployment and so mechanisms such as feature flags [https://martinfowler.com/articles/feature-toggles.html] are introduced in order to decouple deployment of code changes from releases of customer functionality. This has numerous benefits for both sides, since the ability to release is no longer dependent on the ability to deploy and vice versa.

A platform supporting a digital product which is available to customers any time of day, seven days a week also has to have the ability to tolerate failure of infrastructure or individual components without failure of the product itself. This is commonly referred to as high-availability [https://en.wikipedia.org/wiki/High\_availability] and is generally achieved by running several instances of the same services spread across geographical locations, so that the application can keep working even when an entire data center is lost. This horizontal replication is also used for scaling the application with customer demand [https://en.wikipedia.org/wiki/Autoscaling].

An important consideration for a product platform is the cost of building it and the ongoing maintenance involved, including infrastructure provisioning, maintenance (security patches) and scaling capacity to support the products that need to run on it. The effort involved should not be under-estimated [http://www.winnou.com/saas.pdf] and is the motivation behind most digital product now running on cloud infrastructure. We will discuss the reasons for this cost reduction in more detail in sections \ref{sec:cost} and \ref{sec:automation}.

Finally, a product platform needs to include features supporting the constraints of running digital products in safety-critical or regulated environments, e.g. medicine or finance. This means supporting information security features (e.g. secure secrets management for credentials), role based access control with good identity and access management [http://csrc.nist.gov/rbac/sandhu96.pdf] and full audit trail and other controls necessary to reduce the risks involved in all of the activities described above.

There has been a clear trend over the past two decades of automating as much of the above activities and treating them as another part of the product itself. This is the core proposition of cloud services and a key concept of the DevOps movement [https://martinfowler.com/bliki/DevOpsCulture.html].

@@

The goal of Agile software delivery methodologies like Lean [Lean Software Development: An Agile Toolkit - Mary Poppendieck] is to eliminate all waste from the process, thus increasing throughput. Focusing on the platform, the waste can be in, for example, being impeded by platform downtime, working around (missing) platform features, repeatedly performing manual tasks or coordinating with other teams to perform tasks (we will discuss the coordination costs in the next section in more detail).

Deployment of code should be automated and the platform should support practices that reduce or eliminate the risks involved (e.g. techniques like rolling updates [], blue-green deployments [https://martinfowler.com/bliki/BlueGreenDeployment.html], canary releases [https://martinfowler.com/bliki/CanaryRelease.html], etc.). Regression testing should also be automated and be reliable enough to make it virtually impossible to deploy a broken application. This automation ultimately enables continuous delivery of customer value as an overarching goal to gain the ability to rapidly improve the product based on real customer behaviour [https://martinfowler.com/bliki/ContinuousDelivery.html].

The platform also needs to provide good support for the chosen architecture of the application itself. The currently accepted best practice for medium to large scale systems is Microservices Architecture [https://martinfowler.com/articles/microservices.html], which provides numerous benefits for modularity and scalability of the overall solution, but comes with its own set of difficulties in managing the orchestration of a large number of microservices that need to work together as a system. The platform should support this approach with service discovery and configuration management features and reduce the overhead that managing microservice orchestration generally involves.

In the long term, the platform should be flexible enough to support safe technical innovation in all directions and enable practices like evolutionary architecture [https://www.thoughtworks.com/insights/blog/microservices-evolutionary-architecture] and continuous improvement [https://en.wikipedia.org/wiki/Kaizen] even for the infrastructure. We will refer to these features using the term \textit{Agile Infrastructure}. Having the ability to evolve the infrastructure design eliminates the need for large up-front planning and enables ongoing innovation in a way that is similar to Agile software delivery.

@@

The overall cost of building and operating a shared product platform is large enough to pose a serious risk to the business. Typically, when things do not go well, this leads to resistance to cut the losses - the sunk cost fallacy [https://youarenotsosmart.com/2011/03/25/the-sunk-cost-fallacy/]. Instead of trying again, or exploring different solutions, the attitude turns into “we will make it work” and results in products being built on a mediocre platform with reliability and scaling issues, increasing risk of preventable outages and slowing down product delivery. This is the final, often unrecognised, cost of a shared product platform - the missed opportunity cost of not delivering as much customer value as possible.

>>>>

Starting with the software itself, it is crucial that it is thoroughly tested before any part of it can be considered done. Testing is now typically largely automated~\cite{TestPyra22:online}, although manual exploratory testing is also strongly advised~\cite{Explorat86:online}. Testing activities fit on a scale of the level of integration of all the components involved, from testing each in isolation (e.g. unit testing) to fully integrated testing of the run-time behaviour of the entire application stack.

Typically, this is supported by a number of different environments~\cite{Professi40:online} starting from the lower ones, where external dependencies of services are stubbed~\cite{TestStub82:online} and moving all the way up to the production environment which serves customers. Traditionally, these environments are named, long-lived, and are often reused for the purposes of various different testing activities (e.g. functional, performance, and penetration testing).

In order to get applications deployed, configured and started in these environments a process of deployment activities is followed, ranging from completely manual to fully automated. There are several different methodologies addressing the change-related risk of deploying to the production environment, ranging from large regular deployments done every few months all the way to Continuous Delivery and Continuous Deployment~\cite{WhatisCo44:online}.

@@

Typically, when following the continuous deployment method, businesses do not want customers to immediately see the change caused by every deployment and so mechanisms such as feature flags~\cite{FeatureT95:online} are introduced in order to decouple deployment of code changes from releases of customer functionality. This has numerous benefits for both sides, since the ability to release is no longer dependent on the ability to deploy and vice versa.

A platform supporting a digital product which is available to customers any time of day, seven days a week also has to have the ability to tolerate failure of infrastructure or individual components without failure of the product itself. This is commonly referred to as high-availability~\cite{HighAvai70:online} and is generally achieved by running several instances of the same services spread across geographical locations, so that the application can keep working even when an entire data center is lost. This horizontal replication is also used for scaling the application with customer demand~\cite{WhatIsAm85:online}.

An important consideration for a product platform is the cost of building it and the ongoing maintenance involved, including infrastructure provisioning, maintenance (security patches) and scaling capacity to support the products that need to run on it. The effort involved should not be under-estimated~\cite{saaspdf24:online} and is the motivation behind most digital product now running on cloud infrastructure. We will discuss the reasons for this cost reduction in more detail in sections \ref{sec:cost} and \ref{sec:automation}.

Finally, a product platform needs to include features supporting the constraints of running digital products in safety-critical or regulated environments, e.g. medicine or finance. This means supporting information security features (e.g. secure secrets management for credentials), role based access control with good identity and access management~\cite{RoleBase56:online} and full audit trail and other controls necessary to reduce the risks involved in all of the activities described above.

There has been a clear trend over the past two decades of automating as much of the above activities and treating them as another part of the product itself. This is the core proposition of cloud services and a key concept of the DevOps movement~\cite{DevOpsCu34:online}.

@@

The goal of Agile software delivery methodologies like Lean~\cite{LeanSoft20:online} is to eliminate all waste from the process, thus increasing throughput. Focusing on the platform, the waste can be in, for example, being impeded by platform downtime, working around (missing) platform features, repeatedly performing manual tasks or coordinating with other teams to perform tasks (we will discuss the coordination costs in the next section in more detail).

Deployment of code should be automated and the platform should support practices that reduce or eliminate the risks involved (e.g. techniques like rolling updates~\cite{Performi30:online}, blue-green deployments~\cite{BlueGree26:online}, canary releases~\cite{CanaryRe28:online}, etc.). Regression testing should also be automated and be reliable enough to make it virtually impossible to deploy a broken application. This automation ultimately enables continuous delivery of customer value as an overarching goal to gain the ability to rapidly improve the product based on real customer behaviour~\cite{Continuo59:online}.

The platform also needs to provide good support for the chosen architecture of the application itself. The currently accepted best practice for medium to large scale systems is Microservices Architecture~\cite{Microser99:online}, which provides numerous benefits for modularity and scalability of the overall solution, but comes with its own set of difficulties in managing the orchestration of a large number of microservices that need to work together as a system. The platform should support this approach with service discovery and configuration management features and reduce the overhead that managing microservice orchestration generally involves.

In the long term, the platform should be flexible enough to support safe technical innovation in all directions and enable practices like evolutionary architecture~\cite{Microser21:online} and continuous improvement~\cite{KaizenTh68:online} even for the infrastructure. We will refer to these features using the term \textit{Agile Infrastructure}. Having the ability to evolve the infrastructure design eliminates the need for large up-front planning and enables ongoing innovation in a way that is similar to Agile software delivery.

@@

The overall cost of building and operating a shared product platform is large enough to pose a serious risk to the business. Typically, when things do not go well, this leads to resistance to cut the losses - the sunk cost fallacy~\cite{TheSunkC23:online}. Instead of trying again, or exploring different solutions, the attitude turns into “we will make it work” and results in products being built on a mediocre platform with reliability and scaling issues, increasing risk of preventable outages and slowing down product delivery. This is the final, often unrecognised, cost of a shared product platform - the missed opportunity cost of not delivering as much customer value as possible.
